Microphysics and astrophysics inspired priors for neutron star mergers

### 2025-02-12 paper.tex
text added to methods section.

\documentclass[twocolumn]{aastex631}
\usepackage{siunitx}

@@

\author{Christian Ecker}
\affiliation{Institut f\"ur Theoretische Physik, Max-von-Laue-Strasse 1, 60438 Frankfurt, Germany}
\affiliation{Frankfurt Institute for Advanced Studies, Ruth-Moufang-Strasse 1, 60438 Frankfurt, Germany}

@@

>>>>

\documentclass[twocolumn]{aastex631}
\usepackage{siunitx}
\usepackage{amsmath}

@@

	\author{Christian Ecker}
	\affiliation{Institut f\"ur Theoretische Physik, Max-von-Laue-Strasse 1, 60438 Frankfurt, Germany}

@@

	Our setup closely follows that of~~\citet{Altiparmak:2022}, which we briefly review here. 
	We begin by constructing a large set of EOSs by stitching together various components.
	At the lowest densities ($n<0.5,n_s$), we adopt the Baym-Pethick-Sutherland (BPS) model~\citep{Baym71} for the neutron-star crust.
	In the range $0.5,n_s < n < 1.1,n_s$, we randomly sample polytropes to span the range between the softest and stiffest EOSs from~\citet{Hebeler:2013nza}.
	At high densities ($n \gtrsim 40,n_s$), corresponding to a baryon chemical potential of $\mu = 2.6,{\rm GeV}$, we impose the perturbative QCD constraint from~\citet{Fraga2014} on the pressure $p(X, \mu)$ of cold quark matter, with the renormalization scale parameter $X$ sampled uniformly in $[1,4]$.

	For the intermediate density range ($1.1,n_s < n \lesssim 40,n_s$), we use the parametrization method of~\citet{Annala2019}, which models the sound speed as a function of the chemical potential, $c_s^2(\mu)$, using piecewise-linear segments:
	\begin{equation} \label{eq:cs2}
		c_s^2(\mu) = \frac{\left(\mu_{i+1}-\mu \right) c_{s,i}^2 + \left(\mu - \mu_i \right) c_{s,i+1}^2}{\mu_{i+1}-\mu_i}~, 
	\end{equation}
	where $\mu_i$ and $c_{s,i}^2$ are parameters defining the $i$-th segment in the range $\mu_i \leq \mu \leq \mu_{i+1}$.
	Throughout this work, we adopt natural units where $c=G=1$.

	The number density follows as 
	\begin{equation} \label{eq:n}
		n(\mu) = n_1 \exp \left({\int_{\mu_1}^\mu \frac{d\mu^\prime}{\mu^\prime c_s^2(\mu^\prime)}}\right)~, 
	\end{equation} where $n_1 = 1.1,n_s$, and $\mu_1 = \mu(n_1)$ is set by the corresponding polytropic EOS.
	The pressure is then obtained via
	\begin{equation} \label{eq:p}
		p(\mu) = p_1 + \int_{\mu_1}^\mu d\mu^\prime , n(\mu^\prime)~,
	\end{equation}
	where the integration constant $p_1$ matches the pressure of the polytrope at $n = n_1$.
	We numerically integrate Eq.~\eqref{eq:p} using seven segments for $c_s^2(\mu)$.

	Using this framework, we generate $XXX$ EOSs by randomly selecting the maximum sound speed $c_{s,{\rm max}}^2 \in [0,1]$ and uniformly sampling the free parameters $\mu_i \in [\mu_1, \mu_{N+1}]$ (where $\mu_{N+1} = 2.6\rm GeV$) and $c_{s,i}^2 \in [0, c_{s,{\rm max}}^2]$. These EOSs are consistent with nuclear theory and perturbative QCD uncertainties.

	To incorporate astrophysical constraints, we solve the Tolman-Oppenheimer-Volkoff (TOV) equations for each EOS and retain only those satisfying the mass measurements of J0348+0432~\citep{Antoniadis2013} ($M = 2.01\pm 0.04,M_{\odot}$) and J0740+6620~\citep{Cromartie2019, Fonseca2021} ($M = 2.08 \pm 0.07,M_{\odot}$), discarding EOSs with a maximum mass $M_{\rm TOV} < 2.0,M_{\odot}$. We also impose NICER radius constraints from J0740+6620~\citep{Miller2021, Riley2021} and J0030+0451~\citep{Riley2019, MCMiller2019b}, rejecting EOSs with $R < 10.75$ km at $M = 2.0,M_{\odot}$ or $R < 10.8$ km at $M = 1.1,M_{\odot}$.

	In the neutron star binary case, we impose an upper bound on the binary tidal deformability from GW170817, requiring $\tilde{\Lambda} < 720$ under low-spin priors~\citep{Abbott2018a}. The tidal deformability is computed as
	\begin{equation} 
		\tilde{\Lambda}_{\rm NSNS} = \frac{16}{13} \frac{ (12M_2 + M_1) M_1^4 \Lambda_1 + (12M_1 + M_2) M_2^4 \Lambda_2}{(M_1 + M_2)^5}~,
	\end{equation}
	where $M_i$, $R_i$, and $\Lambda_i = \frac{2}{3} k_2 \left( R_i/M_i \right)^5$ ($i=1,2$) denote the component masses, radii, and tidal deformabilities, with $k_2$ being the second tidal Love number.
	EOSs violating $\tilde{\Lambda} < 720$ for a chirp mass $\mathcal{M}_{\rm chirp} = (M_1 M_2)^{3/5} (M_1+M_2)^{-1/5} = 1.186,M_\odot$ and mass ratio $q = M_2/M_1 > 0.73$ are discarded.

	In the mixed binary case, i.e., where one binary component is a neutron star and the other one a black hole, the formula for the binary tidal deformability simplifies, because black holes have zero tidal deformability $\Lambda_{\rm BH}=0$.
	The binary tidal deformability in this case becomes
	\begin{equation} 
		\tilde{\Lambda}_{\rm NSBH} = \frac{16}{13} \frac{ (12M_{\rm BH} + M_{\rm NS}) M_{\rm NS}^4 \Lambda_{\rm NS}}{(M_{\rm NS} + M_{\rm BH})^5}~,
	\end{equation}
	where $M_{\rm NS}$ and $\Lambda_{\rm NS}$ are the nuetron star mass and tidal defromability, respectively, and $M_{\rm BH}$ the black hole mass. The expressions for the chirp mass and mass ratio remain intact, with the understanding that $M_1=M_{\rm NS}$ and $M_2=M_{\rm BH}$. 
	%Applying these observational constraints reduces our initial set of $8 \times 10^6$ EOSs to approximately $10^6$ viable models, forming the basis for the results in the next section.

	%While an upper bound on $M_{\rm TOV}$ has been proposed ($M_{\rm TOV} \lesssim 2.16^{+0.17}{-0.15} M{\odot}$, see \citet{Rezzolla2017} and also \citet{Margalit2017, Ruiz2017, Shibata2019, Nathanail2021}), this estimate relies on model assumptions about the kilonova associated with GRB170817A from GW170817. Given these uncertainties, we do not impose this bound in our main analysis but explore its impact in Appendix~\ref{sec:appendix_a}.
